---
title: "data processing for Obermayer et al."
output: 
  html_document:
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE)
library(sctransform)
library(Seurat)
library(SeuratDisk)
library(ggplot2)
library(cowplot)
library(ggrepel)
library(dplyr)
library(scRepertoire)
library(tidyr)
library(DESeq2)
options(future.globals.maxSize = 10 * 1024 * 1024^2)
```

we follow the [Seurat vignette](https://satijalab.org/seurat/v4.0/reference_mapping.html) and use their PBMC data as reference

```{r get_reference, fig.width=6,fig.height=5}
reference <- LoadH5Seurat(file.path('..','data','seurat','pbmc_multimodal.h5seurat'))
```

get marker genes from this reference

```{r get_reference_markers}
Idents(reference) <- 'celltype.l2'
level2.markers.RNA <- FindAllMarkers(reference,assay='SCT', logfc.threshold=1, min.pct=.25,only.pos=TRUE)
write.csv(level2.markers.RNA, file.path('..','data','seurat','level2_markers_RNA.csv'))
```

load PBMC data and take all cells with less than 10% mitochondrial RNA and between 250 and 5000 genes

- D36prePBMC
- D36postPBMC
- R36d90PBMC
- D26prePBMC
- D26postPBMC
- R26d90PBMC
- R26d180PBMC
- D10prePBMC
- D10postPBMC
- R10d180PBMC
- D16prePBMC
- D16postPBMC
- R16d90PBMC
- R16d180PBMC
- D29postPBMC
- R29d90PBMC
- R29d180PBMC
- D31postR24d30PBMC (only D31post)
- R31d90R31d180PBMC

```{r get_data}
libraries <- c("D36prePBMC","D36postPBMC","R36d90PBMC",
               "D26prePBMC","D26postPBMC","R26d90PBMC","R26d180PBMC",
               "D10prePBMC","D10postPBMC","R10d180PBMC",
               "D16prePBMC","D16postPBMC","R16d90PBMC","R16d180PBMC",
               "D29postPBMC","R29d90PBMC","R29d180PBMC",
               "D31postR24d30PBMC","R31d90R31d180PBMC")

pbmc <- list()
for (lib in libraries) {
  dge <- Read10X_h5(file.path('..','data','cellranger',paste0(lib,'_filtered_counts.h5')))
  sobj <- CreateSeuratObject(counts = dge[['Gene Expression']], project=lib)
  sobj[['pct.mito']] <- PercentageFeatureSet(sobj, pattern='^MT-')
  sobj <- subset(sobj, (pct.mito < 10) & (nFeature_RNA <= 5000) & (nFeature_RNA >= 250))
  cells <- Cells(sobj)
  hts <- grepl('Hashtag[0-9]',row.names(dge[['Antibody Capture']]))
  sobj[["ADT"]] <- CreateAssayObject(counts = dge[['Antibody Capture']][!hts,cells,drop=FALSE])
  sobj <- NormalizeData(sobj, normalization.method='CLR', assay='ADT', verbose=FALSE, margin=2)
  if (lib=='D31postR24d30PBMC') {
    hts <- c('Hashtag3','Hashtag5')
    sobj_HTO <- CreateSeuratObject(counts = dge[['Antibody Capture']][hts,cells],drop=FALSE,assay='HTO')
    sobj_HTO <- NormalizeData(sobj_HTO, normalization.method='CLR', assay='HTO', verbose=FALSE)
    sobj_HTO <- HTODemux(sobj_HTO, assay='HTO', verbose=FALSE)
    vireo <- read.table(file.path('..','data','vireo',paste0(lib,'_donor_ids.tsv')),
                        header=1,row.names=1,as.is=TRUE)
    sobj_HTO <- AddMetaData(sobj_HTO, vireo)
    pbmc[['D31postPBMC']] <- subset(sobj, cells=Cells(sobj_HTO)[(sobj_HTO$donor_id=='donor0') & (sobj_HTO$hash.ID %in% c('Hashtag3','Negative'))])
  } else if (lib=='R31d90R31d180PBMC') {
    hts <- c('Hashtag3','Hashtag5')
    sobj_HTO <- CreateSeuratObject(counts = dge[['Antibody Capture']][hts,cells],drop=FALSE,assay='HTO')
    sobj_HTO <- NormalizeData(sobj_HTO, normalization.method='CLR', assay='HTO', verbose=FALSE)
    sobj_HTO <- HTODemux(sobj_HTO, assay='HTO', verbose=FALSE)
    pbmc[['R31d90PBMC']] <- subset(sobj, cells=Cells(sobj_HTO)[sobj_HTO$hash.ID=='Hashtag3'])
    pbmc[['R31d180PBMC']] <- subset(sobj, cells=Cells(sobj_HTO)[sobj_HTO$hash.ID=='Hashtag5'])
  } else {
    pbmc[[lib]] <- sobj
  }
}
```

get some QC metrics

```{r QC,fig.width=10,fig.height=5}
md <- do.call(rbind, lapply(names(pbmc), function(x) pbmc[[x]]@meta.data[,c('nCount_RNA','nFeature_RNA','pct.mito')] %>%
                              dplyr::mutate(orig.ident=x))) %>%
  dplyr::group_by(orig.ident) %>%
  dplyr::mutate(orig.ident=paste0(orig.ident,' (n=',n(),')')) %>%
  gather(metric,value,-orig.ident)

ggplot(md,aes(x=orig.ident,y=value,fill=orig.ident)) + 
  geom_violin(scale='width') +
  facet_wrap(~metric,scales='free_y') + 
  theme_classic() + 
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
        legend.position='none') + 
  labs(x='',y='')
```

project our cells onto this reference embedding using RNA data; predict cell type labels for the level 1, level 2 and level 3 clustering of the reference

```{r project}
for (sample in names(pbmc)) {
  sobj <- pbmc[[sample]]
  capture.output(sobj <- SCTransform(sobj, verbose = FALSE),file='/dev/null')
  anchors <- FindTransferAnchors(
    reference = reference,
    query = sobj,
    normalization.method = "SCT",
    reference.reduction = "spca",
    dims = 1:50,
    verbose=FALSE)
  pbmc[[sample]] <- MapQuery(
    anchorset = anchors,
    query = sobj,
    reference = reference,
    refdata = list(
      celltype.l1 = "celltype.l1",
      celltype.l2 = "celltype.l2",
      celltype.l3 = "celltype.l3",
      predicted_ADT = "ADT"
    ),
    reference.reduction = "spca", 
    reduction.model = "wnn.umap",
    verbose=FALSE)
}
```

identify and remove putative doublets (shown in pink), and cells with prediction score < .75

```{r doublets,fig.width=10,fig.height=12}
source('my_doubletFinder.R')
pbmc <- lapply(pbmc, function(sobj) {
  DefaultAssay(sobj) <- 'RNA'
  Idents(sobj) <- 'predicted.celltype.l1'
  homotypic.prop <- modelHomotypic(Idents(sobj))
  nExp_poi <- round(0.05*nrow(sobj@meta.data)) 
  nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
  tmp <- my_doubletFinder(sobj, PCs = 1:10, pN = 0.25, pK = 0.09, nExp = nExp_poi, reuse.pANN = FALSE)
  tmp <- my_doubletFinder(tmp, PCs = 1:10, pN = 0.25, pK = 0.09, nExp = nExp_poi.adj, reuse.pANN = "pANN")
  AddMetaData(sobj, tmp@meta.data[,'DF.classifications',drop=FALSE])
})

do.call(rbind,lapply(pbmc, function(sobj) table(sobj$DF.classifications))) %>%
  DT::datatable()

plots <- lapply(names(pbmc), function(sample)
  DimPlot(pbmc[[sample]], group.by='DF.classifications',
          cols=c('Singlet'='gray',
                 'Doublet'='deeppink')) +
    theme_void() + 
    labs(title=sample) + 
    theme(legend.position='none'))

plot_grid(plotlist=plots,ncol=5,align='vh')

pbmc.sub <- lapply(pbmc, function(x) subset(x, subset=(DF.classifications=='Singlet') & (predicted.celltype.l1.score >= .75)))
```

merge all datasets

```{r merge}
pbmc.all <- merge(pbmc.sub[[1]],pbmc.sub[2:length(pbmc.sub)],
                  add.cell.ids=names(pbmc.sub))

umap <- do.call(rbind, lapply(seq(1,length(pbmc.sub)), function(i) {
  emb <- Embeddings(pbmc.sub[[i]], reduction='ref.umap')
  row.names(emb) <- paste0(names(pbmc.sub)[[i]],'_',row.names(emb))
  emb })
)
pbmc.all[["umap"]] <- CreateDimReducObject(embeddings = umap, key = "UMAP_", assay = DefaultAssay(pbmc.all))
pbmc.all$sample <- gsub('_.*','',Cells(pbmc.all))
pbmc.all$donor <- gsub('[prepostd]*[0-9]*PBMC','',pbmc.all$sample)
pbmc.all$donor_type <- ifelse(grepl('^D',pbmc.all$sample),'donor','recipient')
```

add clonotype information

```{r add clonotypes}
bcs <- sapply(unique(pbmc.all$sample), function(x) gsub('.*_([ACGT]*-[0-9]*)','\\1',Cells(pbmc.all)[pbmc.all$sample==x]))
  
tcr_list <- list()
for (lib in unique(pbmc.all$orig.ident)) {
  infile <- file.path('..','data','cellranger',paste0(lib,'_filtered_contigs.csv'))
  if (file.exists(infile)) {
    tcr <- read.csv(infile,
                    stringsAsFactors=FALSE) %>%
      dplyr::mutate(is_cell=is_cell=='true',
                    high_confidence=high_confidence=='true',
                    full_length=full_length=='true',
                    productive=productive=='true')
    for (sample in unique(pbmc.all$sample[pbmc.all$orig.ident==lib])) {
      tcr_list[[sample]] <- tcr %>% dplyr::filter(barcode %in% bcs[[sample]])
    }
  }
}
  
tcr_combined <- sapply(which(sapply(tcr_list, function(x) !is.null(x))),
                       function(i) combineTCR(tcr_list[[i]],
                                              samples=c(names(tcr_list)[[i]]),
                                              ID=c(names(tcr_list)[[i]]),
                                              cells='T-AB',
                                              filterMulti=TRUE))
names(tcr_combined) <- names(tcr_list)[sapply(tcr_list, function(x) !is.null(x))]
  
saveRDS(tcr_combined,file.path('..','data','seurat','tcr_combined_all.rds'))
```

add TCR information to Seurat objects

```{r add_clonotypes,fig.width=12,fig.height=5}
pbmc.all <- RenameCells(pbmc.all, add.cell.id <- pbmc.all$sample)

tmp <- lapply(tcr_combined, function(x) x[x$barcode %in% Cells(pbmc.all),])
pbmc.all <- combineExpression(tmp, pbmc.all,
                              cloneCall="aa")
```

```{r save}
saveRDS(pbmc.all,file.path('..','data','seurat','pbmc_all.rds'))
```

```{r sessionInfo}
sessionInfo()
```

